---
title: "Final Project: Checkpoint 1"
subtitle: "STAT 244"
author: "Eonbi Choi, Yerim Oh"
output:
  pdf_document:
    fig_height: 1.5
    fig_width: 4
    highlight: zenburn
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(readr)
library(purrr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(mosaic)
library(gmodels)
library(Sleuth3)
```

### 1. Review the \href{https://docs.google.com/document/d/11G6wyZeNyfobAQm-yCaBj0d8C3wCuxCmKqwaErpTMcQ/edit?usp=sharing}{\textcolor{blue}{Final Project Description document}}

### 2. Describe your data set in 1 - 2 sentences.

The Formula 1 data used in this study are obtained from the f1dataR R package that accesses Formula 1 data via the FastF1 Python library. The dataset includes lap-by-lap session data from the 2024 Miami Grand Prix and comprise 1111 laps and 32 variables. These variables include driver details, lap times, pit in/out times, tire information, and track status.

### 3. What is the source of your data set? Include a link to where you got it here.

**Package ‘f1dataR’**

-   **Description**: <https://cran.r-project.org/web/packages/f1dataR/f1dataR.pdf>

-   **Data sources**: Obtain Formula 1 data via the unofficial API <https://www.formula1.com/en/timing/f1-live> via the 'fastf1' 'Python' library <https://docs.fastf1.dev/>.

### 4. Load your data set using the code chunk below.

#### Load packages needed for the data

```{r, warning= FALSE, message=FALSE, fig.align='left'}
# Install f1dataR package
#install.packages("f1dataR")
#setup_fastf1()     # Run during initial setup
library(f1dataR)
```

```{r, warning= FALSE, message=FALSE, fig.align='left'}
#miami2024 <- load_session_laps(season = 2024, session = "R", round = 6)
```

```{r}
load("lap_dat.Rdata")

# rearrange data
miami2024 <- lap_dat %>%
  select(lap_time, lap_number, compound, tyre_life) %>%
  mutate(compound = as.factor(compound),
         pit_in = ifelse(is.nan(lap_dat$pit_in_time), 0, 1))
head(miami2024)
```

-   We transformed the variable `pit_in_time` into a binary variable `pit_in`, where `NaN` values in `pit_in_time` were converted to `0` in `pit_in`, indicating no pit stop, and numeric values were converted to `1`, indicating a pit stop occurred.

### 5. How many rows does your data set have? There should be at least 30 rows.

```{r}
nrow(miami2024)
```

There are 1111 rows

### 6. How many quantitative variables are in your data set? There should be at least 3. List them here and briefly describe what they represent **and** their units.

There are 3 quantitative variables

-   `lap_time`: recorded time to complete a lap

-   `lap_number`: lap number from which the telemetry data was recorded

-   `tyre_life`: number of laps completed on a set of tires

### 7. How many categorical variables does your data set have? There should be at least 1. List the categorical variable(s) here along with their corresponding categories.

There are two categorical variables

-   `compound`: type of tire used

-   `pit_in`: whether a driver made a pit stop during a lap (binary: 0 = no pit stop, 1 = pit stop occured)
